Biogeography of crop progenitors and wild plant resources over in terminal Pleistocene and Early Holocene boundary West Asia, 14.7--8.3 ka

### 2025-01-05 analysis/paper.tex
Add discussion

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Biogeography of crop progenitors and wild plant resources over the Pleistocene/Holocene boundary in West Asia, 14.7--8.3 ka},
  pdfauthor={Joe Roe; Amaia Arranz-Otaegui},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Biogeography of crop progenitors and wild plant resources over
the Pleistocene/Holocene boundary in West Asia, 14.7--8.3 ka}
\author[1,2]{Joe Roe%
\corref{cor1}%
}
 \ead{joeroe@hey.com} 
\author[3]{Amaia Arranz-Otaegui%
%
}

@@

\begin{abstract}
This paper presents the first comprehensive reconstructions of the
palaeodistributions of 56 plant species found regularly in association
with early agricultural archaeological sites in West Asia, including the
progenitors of the first crops. We used machine learning to train an
ecological niche model of each species based on its present-day
distribution in relation to climate and environmental variables.
Predictions of the potential ranges of these species at key stages of
the Pleistocene--Holocene transition could then be derived from these
models using hindcast data from palaeoclimate simulations. Species
ranges are predicted to have been on average {[}X\%{]}
{[}larger\textbar smaller{]} in the Early Holocene compared to present
conditions, but did {[}not{]} change as drastically between the Last
Glacial Maximum and the Early Holocene. However, the modelled ranges do
not predict the observed occurrence of these species on archaeological
sites with any degree of accuracy. The regional ubiquity of species in
the archaeological record is {[}not{]} correlated with the predicted
size of its range and the diversity of archaeobotanical assemblages is
{[}not{]} correlated with the predicted diversity of its environs. This
indicates that trends in taxonomic composition of the archaeobotanical
record is {[}not{]} likely to have been influenced by environmental
change and species turnover, in addition to human economic choices.
\end{abstract}

@@

\begin{itemize}
\tightlist
\item[$\boxtimes$]
  Finish modelling
\item[$\square$]
  First draft:

@@

    \begin{itemize}
    \tightlist
    \item[$\square$]
      Biogeography
    \item[$\boxtimes$]
      ENM
    \end{itemize}
  \item[$\boxtimes$]
    Methods \& Materials
  \item[$\square$]
    Results

    \begin{itemize}
    \tightlist
    \item[$\square$]
      Model assessment
    \end{itemize}
  \item[$\square$]
    Discussion

    \begin{itemize}
    \tightlist
    \item[$\square$]
      General trends
    \item[$\square$]
      Case studies
    \end{itemize}
  \item[$\square$]
    Conclusion
  \end{itemize}
\item[$\square$]
  Figures

  \begin{itemize}
  \tightlist
  \item[$\square$]
    tbl-occ-count (needs pdf cleanup)
  \item[$\square$]
    tbl-predictors
  \item[$\square$]
    tbl-climate-periods
  \end{itemize}
\item[$\square$]
  Clean up package code
\item[$\square$]
  Appendix with all hindcast predictions
\item[$\square$]
  References \& copyediting
\item[$\square$]
  Final proofread
\end{itemize}

@@

  \begin{itemize}
  \tightlist
  \item
    More than the ``founder crops'';
  \item
    More than food
  \item
    (In archaebot., not all intentionally collected)
  \item
    Regionally and temporally diverse
  \item
    \ldots so we model lots of species!
  \end{itemize}
\item
  Regional ecological reconstructions generally rely on the `expert
  interpolation' (or what do they call it with isoscapes?) method

@@

We began by considering 62 distinct taxa (Table~\ref{tbl-occ-count}) -
all the identifiable species known to be present at more than three
Neolithic (c.~11.7--6.5 ka) sites in West Asia, according to our
previous study of the regional archaeobotanical data
\citep{ArranzOtaeguiRoe2023}. Taxonomic names were resolved to the
canonical form specified in the GBIF Backbone Taxonomy
\citep{GBIFSecretariat2023}. So for example occurrences for
\emph{Bolboschoenus maritimus} also include those recorded under the
older nomenclature \emph{Scirpus maritimus} (see
Table~\ref{tbl-occ-count}). Domestic species meeting our inclusion
criteria were substituted with their wild progenitor(s), where known.

@@

@@

Ecological niche modelling is a classification problem that can be
approached with a wide range of statistical methods. A substantial
literature exists on the relatively performance of these approaches and
their respective parameterisations \citep[reviewed
in][]{ValaviEtAl2022}. Random Forest, a widely-used machine learning
algorithm, is amongst the best performing approaches for presence-only
species distribution models, providing it is appropriately parameterised
to account for the class imbalance between presence and background
samples \citep{ValaviEtAl2021, ValaviEtAl2022}. For our application, it
also has the advantage of requiring little to no manual parameter tuning
to achieve good predictive results, which makes it easier to model a
larger numbers of taxa.

@@

We assessed the predictive performance of the fitted niche models in the
contemporary environment based on the reserved test partition. Model
accuracy (proportion of correctly classified presence and background
samples) ranged between 72\% and 98\%, with an average of 91\%.
Sensitivity (proportion of correctly classified presence samples) ranged
between 80\% and 100\%, with an average of 93\%. The area under the
models' receiver operating characteristic curves (ROC-AUC) was in the
range of 0.973±0.058 Model sensitivity is loosely correlated with the
number of occurrences available for training
(Figure~\ref{fig-model-sensitivity-vs-n_occ}), with the worst-performing
models all having less than 300 recorded occurrences: \emph{Buglossoides
tenuiflora}, \emph{Cephalaria syriaca}, \emph{Gypsophila pilosa},
\emph{Lathyrus sativus}, \emph{Triticum turgidum dicoccum}. Test metrics
and ROC curves for the individual models are included in the appendix.

@@

\begin{itemize}
\tightlist
\item
  General trends:

  \begin{itemize}
  \tightlist
  \item
    Quantified sensitivity of plant ranges to climate change
  \item
    Crop progenitors saw range contractions just before the onset of
    agriculture? (Moreso than other wild resources??)
  \end{itemize}
\item
  Interesting individual case studies:

  \begin{itemize}
  \tightlist
  \item
    Wheat progenitors
  \item
    ???
  \end{itemize}
\end{itemize}

\subsection{Comparison with archaeobotanical
data}\label{comparison-with-archaeobotanical-data}

@@

At the same time, we cannot rule out more substantive reasons for the
discrepency between predicted and observed archaeological occurrences.
The niches of the modelled species could have changed since the Early
Holocene, which would not be captured in a model trained purely on
modern specimens. Human economic choices---mobility, foraging
strategies, cultivation, etc.---could also produce archaeobotanical
assemblages whose composition depart significantly from that of the
surrounding local flora. Further refinement of the methodology for
hindcast palaeoecological niche models, for example using more finely
resolved palaeoclimate sequences \citep{KargerEtAl2023}, hyperparameter
tuning to avoid overfitting, and improved archaeological datasets, would
help disentangle these potential explanations.

@@

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Biogeography of crop progenitors and wild plant resources over in terminal Pleistocene and Early Holocene boundary West Asia, 14.7--8.3 ka},
  pdfauthor={Joe Roe; Amaia Arranz-Otaegui},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Biogeography of crop progenitors and wild plant resources over in
terminal Pleistocene and Early Holocene boundary West Asia, 14.7--8.3
ka}
\author[1,2]{Joe Roe%
\corref{cor1}%
}
 \ead{joeroe@hey.com} 
\author[3]{Amaia Arranz-Otaegui%
%
}

@@

\begin{abstract}
This paper presents the first continuous, spatially-explicit
reconstructions of the palaeodistributions of 56 plant species found
regularly in association with early agricultural archaeological sites in
West Asia, including the progenitors of the first crops. We used machine
learning to train an ecological niche model of each species based on its
present-day distribution in relation to climate and environmental
variables. Predictions of the potential ranges of these species at key
stages of the Pleistocene--Holocene transition could then be derived
from these models using downsampled data from palaeoclimate simulations.
Species ranges are predicted to have been on significantly smaller in
the Early Holocene compared to present conditions, with many species
that are found throughout the region's `hilly flanks' today indicated to
have much more restricted distributions centered on the Levant, Cyprus
and Western Anatolia. Ranges shrunk by on average c.~10\% from the
terminal Pleistocene to the Early Holocene, with a further c.~10\%
reduction for the duration of the Younger Dryas. However, modelled
ranges do not reliably predict the observed occurrence of specific
species at archaeological sites. {[}\ldots{]} The regional ubiquity of
species in the archaeological record is {[}not{]} correlated with the
predicted size of its range and the diversity of archaeobotanical
assemblages is {[}not{]} correlated with the predicted diversity of its
environs. This indicates that trends in taxonomic composition of the
archaeobotanical record is {[}not{]} likely to have been influenced by
environmental change and species turnover, in addition to human economic
choices.
\end{abstract}

@@

\begin{itemize}
\tightlist
\item[$\boxtimes$]
  Stop modelling
\item[$\square$]
  First draft:

@@

    \begin{itemize}
    \tightlist
    \item[$\square$]
      Biogeography
    \item[$\boxtimes$]
      ENM
    \end{itemize}
  \item[$\boxtimes$]
    Methods \& Materials
  \item[$\boxtimes$]
    Results
  \item[$\square$]
    Discussion

    \begin{itemize}
    \tightlist
    \item[$\square$]
      General trends
    \item[$\square$]
      Case studies
    \item[$\boxtimes$]
      Comparison with archaeobot
    \end{itemize}
  \item[$\square$]
    Conclusion
  \end{itemize}
\item[$\square$]
  Figures

  \begin{itemize}
  \tightlist
  \item[$\boxtimes$]
    fig-region
  \item[$\boxtimes$]
    fig-occ
  \item[$\square$]
    tbl-occ-count (needs pdf cleanup)
  \item[$\square$]
    tbl-predictors
  \end{itemize}
\item[$\square$]
  Outstanding TODOs
\item[$\square$]
  Add missing references
\item[$\square$]
  Copyedit
\item[$\square$]
  Appendix with all hindcast predictions (HTML?)
\item[$\square$]
  Clean up package code
\item[$\square$]
  Final proofread
\end{itemize}

@@

  \begin{itemize}
  \tightlist
  \item
    More than the ``founder crops'';
  \item
    More than food
  \item
    (In archaebot., not all intentionally collected)
  \item
    Geographically and temporally diverse
  \item
    \ldots so we model lots of species!
  \end{itemize}
\item
  Regional ecological reconstructions generally rely on the `expert
  interpolation' (or what do they call it with isoscapes?) method

@@

We began by considering 62 distinct taxa (Table~\ref{tbl-occ-count}) -
all the identifiable species known to be present at more than three
Neolithic (c.~11.7--6.5 ka) sites in West Asia, according to our
previous study of the regional archaeobotanical data
\citep{ArranzOtaeguiRoe2023}. Taxonomic names were resolved to the
canonical form specified in the GBIF Backbone Taxonomy
\citep{GBIFSecretariat2023}. So for example occurrences for
\emph{Bolboschoenus maritimus} also include those recorded under the
older nomenclature \emph{Scirpus maritimus} (see
Table~\ref{tbl-occ-count}). Domestic species meeting our inclusion
criteria were substituted with their wild progenitor(s), where
different.

@@

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-occ-map-1.pdf}}

\caption{\label{fig-occ-map}Georeferenced occurrence records from West
Asia for the \texttt{r\ nrow(flora)} modelled taxa
(N=\texttt{r\ nrow(unnest(flora,\ occ))})}

@@

Ecological niche modelling is a classification problem that can be
approached with a wide range of statistical methods. A substantial
literature exists on the relatively performance of these approaches and
their respective parameterisations \citep[reviewed
in][]{ValaviEtAl2022}. Random Forest, a widely-used machine learning
algorithm, is amongst the best performing methods for presence-only
species distribution models, providing it is appropriately parameterised
to account for the class imbalance between presence and background
samples \citep{ValaviEtAl2021, ValaviEtAl2022}. For our application, it
also has the advantage of requiring little to no manual parameter tuning
to achieve good predictive results, which makes it easier to model a
larger numbers of taxa.

@@

We assessed the predictive performance of the fitted niche models in the
contemporary environment based on the reserved test partition. Model
accuracy (proportion of correctly classified presence and background
samples) ranged between 72\% and 98\%, with an average of 91\%.
Sensitivity (proportion of correctly classified presence samples) ranged
between 80\% and 100\%, with an average of 93\%. The area under the
models' receiver operating characteristic curves (ROC-AUC) was in the
range of 0.973±0.058 Model sensitivity is loosely correlated with the
number of occurrences available for training
(Figure~\ref{fig-model-sensitivity-vs-n_occ}), with the worst-performing
models all having less than 300 recorded occurrences: \emph{Buglossoides
tenuiflora}, \emph{Cephalaria syriaca}, \emph{Gypsophila pilosa},
\emph{Lathyrus sativus}, and \emph{Triticum turgidum dicoccum}. Test
metrics and ROC curves for the individual models are included in the
appendix.

@@

\subsection{Reduction in range sizes over the Pleistocene/Holocene
boundary}\label{reduction-in-range-sizes-over-the-pleistoceneholocene-boundary}

\pandocbounded{\includegraphics[keepaspectratio]{paper_files/figure-pdf/fig-range-density-1.pdf}}

\caption{\label{fig-range-density}Distribution of predicted species
ranges by period. Dashed lines indicate the median range.}

Species ranges are predicted to have been signicantly smaller in the
terminal Pleistocene and Early Holocene that under current conditions
Figure~\ref{fig-range-density}, though the magnitude of this change is
likely also to reflect a degree of overfitting in the model (see below).
Fluctuations in modelled range size between the Bølling-Allerød
(14.7--12.9 ka), Younger Dryas (12.9--11.7 ka), and Early Holocene
(11.7--8.3 ka) are more directly comparable. The average range of
modelled species shrank by 12\% over the Pleistocene--Holocene boundary,
with a noticeable dip in the Younger Dryas. This perhaps indicates that
although this period is considered one of climatatic amelioration
globally \citep{JonesEtAl2019}, the colder conditions of the Pleistocene
may have supported more extensive plant-based economies in West Asia
specifically.

Many taxa that occur (or are predicted to occur) across the `hilly
flanks' today---including most crop progenitors---are reconstructed to
have had a significantly more restricted distribution in the terminal
Pleistocene/Early Holocene. These include \emph{Ficus carica} (fig);
\emph{Hordeum} spp. (wild barleys); \emph{Lathyrus aphaca} and \emph{L.
sativus} (both marginally edible legumes); \emph{Triticum aestivum
compactum} (in the N. Levant), \emph{T. monococcum aegilopoides},
\emph{T. durum}, and \emph{Triticum urartu} (but not the other wheat
progenitor, \emph{T. turgidum dicoccum} -- see below); \emph{Aegilops
speltoides}, but not \emph{Aegilops tauschii} (goatgrasses); and
\emph{Vicia} spp. (vetches), including \emph{Vicia faba} (broad beans).
Central Anatolia, Northern Mesopotamia, and the Zagros Mountains in
particular disappear from the predicted ranges of these species, leaving
the Levant and to a lesser extent Western Anatolia Cyprus as refugia.

Our results for the Levant are consistent with the current understanding
of this region as developing early intensive foraging economies {[}the
Natufian culture, @{]} and as a centre of origin of agriculture {[}@{]}.
It should also be noted that the selection of flora for this study
(Table~\ref{tbl-occ-count}) is biased towards Levantine fauna due to the
history of archaeological research in the region, as is the occurrence
dataset used to train the model Figure~\ref{fig-occ-map}. Within the
Levant, retreat from the Badia.

Loss of the Northern Mesopotamia--Southeastern Anatolia region from the
predicted ranges of crop progenitors is interesting in light of the
`golden triangle' hypothesis \citep{KozlowskiAurenche2005}, which puts
this region at the centre of the development of agriculture. Multiple
lines of archaeological evidence have emerged that point away from this
hypothesis and towards a more geographically diverse origin
\citep{Asouti2006, cites}, but it remains the area with some of the
earliest clear evidence of domestication. Comparative genetics also
points to the Northern Mesopotamia region as the centre of diversity of
many crops \citep{cite}. But since these studies are based on modern
genomes, if the wild range of these plants has, as our modelling
suggests, shifted since the Early Holocene, the apparent centre of
diversity may have shifted with them. Our reconstructions are consistent
with the late arrival of intensive plant-based foraging economies in
this region (cf.~the Natufian of the Levant) and more broadly there need
not be a link between the core \emph{wild} range of a plant and the core
area of its domestication. A scenario in which cultivation emerged at
the edges of the ranges of valuable plant resources---as a means of
extend their natural niche---is also plausible.

The near-absence of the Zagros in any predicted ranges is also
surprising, given mounting evidence that domestication took place just
as early in the eastern \emph{Mashriq} as it did in the west
\citep{Braidwood, GanjDarehGoats, HillyFlanksStuff}. We consider that
the most likely explanation for this is that our flora does not include
the species that were most important to plant subsistence in the east.
Archaeobotanical data on Neolithic sites in the Zagros is limited
(compared to the Levant in particular) due to a hiatus in field research
there from the 1980s to early 2010s \citep{cite}. Recent research
\citep{Amaia} indicates \ldots{}

Cyprus and W. Anatolia not considered part of the core zone -- maybe
they should be?

Exceptions to the trend of range reduction include \emph{Cicer
reticulatum} (wild chickpea), which has a relatively stable range
centered on Northern Mesopotamia; and \emph{Triticum turgidum dicoccum}
(wild emmer wheat), which is predicted to have two very restricted
ranges centered around the Black Sea Coast of Anatolia and the Palmyra
basin. In the latter case, neither of these areas are part of the
predicted modern distribution of wild emmer (centered around the
Caucasus and Northern Mesopotamia), but it would be consistent with
archaeological evidence for early cultivation at Tell Abu Hureyra and
other sites in the Upper Euphrates. On the other hand, the model for
\emph{Triticum turgidum dicoccum} has the lowest predictive performance
of all our fitted models.

\subsection{Biogeography of crop
progenitors}\label{biogeography-of-crop-progenitors}

Several crop progenitors reduced in range before cultivation? Reduction
in range from BA to EH very evident in Hordeum, Pistacia atlantica.

The wheat story.

Flax has a very restricted distribution (consistent with low occurrence
in founders paper?). As does Pistacia atlantica, Bolboschoenus maritimus

\emph{Secale cereale} = an Anatolian boy

\subsection{Hindcast models do not predict archaeobotanical
composition}\label{hindcast-models-do-not-predict-archaeobotanical-composition}

@@

At the same time, we cannot rule out more substantive reasons for the
discrepency between predicted and observed archaeological occurrences.
The niches of the modelled species could have changed since the Early
Holocene, which would not be captured in a model trained purely on
modern specimens. Human economic choices---mobility, foraging
strategies, cultivation, etc.---could also produce archaeobotanical
assemblages whose composition depart significantly from that of the
surrounding local flora. Further refinement of the methodology for
hindcast palaeoecological niche models, for example using more finely
resolved palaeoclimate sequences \citep[e.g.][]{KargerEtAl2023},
hyperparameter tuning to avoid overfitting, and improved archaeological
datasets, would help disentangle these potential explanations.

@@

\begin{itemize}
\tightlist
\item
  We present the first continuous, spatially explicit models of the
  palaeodistributions of 56 plant species found regularly in association
  with early agricultural archaeological sites in West Asia

  \begin{itemize}
  \tightlist
  \item
    A new line of evidence on archaeoecology
  \item
    Complementary to archaeobot/pollen/etc. because it is independent
    from it
  \item
    All models are wrong\ldots{} but it's easier to see how these are
    wrong than lines on maps
  \end{itemize}
\item
  Modelling at scale using random forest, modern occurrences, and
  hindcasting represents a significant advance in pENM methodology

  \begin{itemize}
  \tightlist
  \item
    Relies on recent open ecological and climatic datasets
  \item
    \ldots open archaeological datasets still lacking!
  \end{itemize}
\item
  First (? - check Yaworsky) attempt to verify hindcast models with
  archaeo. compositional data

  \begin{itemize}
  \tightlist
  \item
    Results not to promising, but this doesn't mean the models are
    useless!
  \item
    Discrepencies suggests several areas for further research and
    methodological development
  \end{itemize}
\end{itemize}
